\TITLE

### 2021-01-27 draft/response-3.tex

\begin{requirement}{\bf Model Orchestration.} \normalfont
  To coordinate the different model predictions and the generation of
  the deep learning forecasts, we need the ability to orchestrate them
  while applying a number of parameters. \Solution To address this
  requirement, the AICov architecture includes the ability to
  integrate parameter sweeps to, for example, identify
  hyperparameters, or integrate different data sets as parameters to
  identify suitable forecast models that can deliver the one which
  fits the best.
\end{requirement}

\begin{requirement}{\bf Compute Resource Mapping.} \normalfont
  For other uses of AICov, we need to integrate a flexible resource
  utilization framework. \Solution To address this requirement, we
  leverage from our earlier work and utilize through Cloudmesh and
  National Institute of Standards and Technology (NIST) Big Data
  Working Group (NBDIF) definitions cloud services into the
  architecture \citep{las-19-nist}. This includes containers but also
  infrastructure services in the cloud. Through the service
  integration, we also provide access to Graphics Processing Units
  (GPUs).
\end{requirement}

\begin{requirement}{\bf Convenient Interfaces.} \normalfont
  As we have a wide variety of users, we need to enable interfaces
  that are used by the various communities. \Solution To address this
  requirement, we provide a number of APIs, REST services and make
  them available via Jupyter Notebooks. In addition, we developed some
  custom widgets for the notebooks that are specifically targeted
  towards data integration, parameter manipulation, and visualization
  of the results.
\end{requirement}

Putting all these requirements together results in an architecture as
shown in Figure \ref{fig:arch}.\\

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.6\columnwidth]{images/arch.pdf}
    \caption{AICov Architecture}
    \label{fig:arch}
\end{figure}

In summary, the AICov architecture allows the inclusion of new data
sources and models. An orchestrator enables modifying parameters and
selecting data sources and models used in the analysis. The framework
can be accessed via the interface layer that includes APIs, REST, and
Jupyter Notebooks. Compute interfaces are provided either by direct
access of GPUs from local resources or through the staging of
automatically generated REST services via our Cloudmesh REST Service
integrator
\citep{cloudmesh-openapi-install,cloudmesh-openapi-benchmark,cloudmesh-manual}.
Additionally, we can also access through abstractions cloud services
that are offered by the various cloud providers and are targeting our
application domain. Different models can be easily integrated into our
framework to verify our forecast or enhance our forecast based on the
users' models and data.

\section{Data for the COVID-19 Analysis}
\label{sec:data}

While the U.S. has seen an overall case fatality rate of 5.14\%, some
U.S. locations have experienced a disproportionate number of deaths
compared to others. This disparity of deaths among different
communities could be attributed to a diversity of local risk factors
and socioeconomic determinants of health that now constitute a very
active area of investigation. Researchers are studying the effects of
different community-specific socioeconomic variables, underlying
health conditions and comorbidities, infrastructure, and systemic
responses.  Given the complexity of such underlying conditions that
can determine individual disease outcomes, where and under what
conditions people get infected by the virus, and then recover or die
take place not in isolation but rather in an interplay of diverse
factors that are characteristic to a local community, which need to be
synthesized. In other words, a basic requirement for effective
modeling of COVID-19 dynamics is that the information regarding a
particular community’s vulnerabilities and strengths pertaining to the
disease must be as comprehensive and integrative as possible.  The
regular reporting of accurate Coronavirus outbreaks data from local
health departments has been understandably difficult, especially in
the areas that are hardest hit.

%\linelabel{comment:level}

We use data on community-specific health and
behavioral risk factors aggregated at the county level from the
large-scale longitudinal CDC Behavioral Risk Factor Surveillance
System (BRFSS) surveys \citep{www-cdc-brfss} and demographic and
socioeconomic data from the U.S. Census Bureau \cite{www-sensus} for each county, along with the cases and
deaths data compiled by the CDC National Center for Health Statistics
(NCHS) \citep{www-cdc-nchs} and integrate them with other data
sources such as the American Hospital Directory \citep{www-ahd}.

\subsection{Data Collection and Processing}

Among the different sources of data on daily cases and deaths data
collected and aggregated at different spatial levels, especially for
the U.S., many are either freely available repositories such as GitHub
or accessible via specifically developed online resources such as the
New York Times~\citep{github-nytimes} and the Johns Hopkins
Coronavirus Resource Center~\citep{www-jh-covid19}.  For this study,
we generated two data sets based on the latter resource upon our
checking for the locations and times at which cumulative counts of
COVID-19 cases and deaths on any given day were no less than those on
the previous day. This led us to compile daily coronavirus cases and
deaths data for 345 U.S. counties from 32 states and the District of
Columbia and matched by five-digit Federal Information Processing
Standards (FIPS) code or county name to dynamic and static variables
from additional data sources.

The static data in this repository was collected from the U.S. Census
Bureau and the Centers for Disease Control and Prevention (CDC). The
different fields cover the domains of behavioral and health risk
factors, hospital capacity, and socioeconomic and demographic
conditions. Data on health outcomes, prevention measures, and
unhealthy behaviors were extracted from the CDC's 500 Cities Local
Data for Better Health program \citep{www-cdc-chronic-data} in the form
of crude or age-adjusted prevalence of conditions such as respiratory
disease, obesity or smoking. These values were aggregated from the
census tract to the county level using the first five digits of the
eleven-digit tract FIPS code. Because a single county may consist of
multiple individual cities, we include the list of all city labels
within each aggregate group to represent a greater metropolitan
area. 110 of such metropolitan areas that had more than 500 reported
cases of COVID-19 by April 15, 2020, were selected for this study.

Notably, a greater metropolitan area such as New York City could be
spread across more than one county, which can lead to the complexity
of aggregating counts and other variables. Time series data for the 5
individual counties (boroughs) in New York City were not available in
the Johns Hopkins CSSE (Center for Systems Science and Engineering)
data set. Rather, the total number of cases and deaths for the entire
city of New York are reported and assigned the FIPS code of New York
County (Manhattan). To ensure consistent geography across the static
and dynamic data, we compute the population-weighted sum or median of
each covariate over the assigned counties.

Demographic variables were gathered from the Census
QuickFacts~\citep{www-census} online resource using an automated web
scraping algorithm and cover relevant areas such as age, race, income,
and population density. Additional socioeconomic variables include the
Gini Index, which measures economic inequality and CDC Social
Vulnerability Index (SVI)~\citep{www-cdc-svi}. The SVI was created to
guide public health officials and disaster response efforts by
identifying the communities across the United States most likely to
need support during a crisis. Census tracts and corresponding counties
are ranked across 15 social factors, which are grouped into four
themes: socioeconomic status, household composition and disability,
minority status and language, and housing and transportation.

Lastly, the total number of general acute care, critical access, and
military hospitals within each county are included in the data. Such
variables included the number of relevant hospitals per county, and
the estimated number of beds (total known bed counts added to the
number of hospitals in the county with missing data times the average
number of beds per hospital in that state) per 1,000 people using the
American Hospital Directory~\citep{www-ahd}.
The list of covariates is presented in Table \ref{tab:risk-factors}.

\begin{table}[!hptb]
\caption{Covariate Risk factor abbreviations as used in this study.}
\label{tab:risk-factors}
\bigskip
\resizebox{0.90\textwidth}{!}{
\centering
\begin{tabular}{ll}
\toprule
Risk Factor & Description \\
\midrule
 NONE & No risk factor used\\
 ARTHRITIS 	  & Percent reported with arthritis \\
 BINGE 		  & Percent reported with binge drinking \\
 BPHIGH 	  & Percent reported with  high blood pressure\\
 BPMED 		  & Percent reported taking blood pressure medication\\
 CANCER 	  & Percent reported as cancer patients\\
 CASTHMA 	  & Percent reported with current asthma\\
 CHD 		  & Percent reported with coronary heart disease\\
 CHECKUP 	  & Percent reported with health checkup\\
 CHOLSCREEN   & Percent reported with cholesterol screening\\
 COPD 		  & Percent reported with chronic obstructive pulmonary disease\\
 CSMOKING 	  & Percent reported as currently smoking\\
 DIABETES 	  & Percent reported with diabetes\\
 ESTBEDS 	  & Number of estimated beds in hospitals\\
 HIGHCHOL 	  & Percent reported with high cholesterol\\
 INSURANCE 	  & Percent reported with insurance\\
 KIDNEY 	  & Percent reported with kidney disease\\
 LPA 		  & Percent reported with no leisure-time physical activity \\
 MHLTH 		  & Percent reported with not good mental health\\
 NBEDS 		  & Number of beds in hospitals \\
 NBEDS/1000   & Number of beds per 1000 people\\
 NHOSP 		  & Number of hospitals\\
 OBESITY 	  & Percent reported with  population that is obese\\
 PHLTH        & Percent reported with physical health issues\\
 PVI 		  & Pandemic vulnerability index \\
 STROKE 	  & Percent reported with stroke\\
 BLACK	  & Percent of Blacks in the population\\
 POVERTY & Percent of poverty in the population\\
 SENIOR  & Percent of seniors in the population\\
 SVI\_MINORITY 	  & Percent of social vulnerability index in the minority population\\
 SVI\_OVERALL 	  & Percent of social vulnerability index in the overall population\\
\bottomrule
\end{tabular}
}
\end{table}

\section{Analysis}

We have conducted a number of analysis efforts to evaluate the validity of our approach. We focus on the following questions:

\begin{question} \normalfont Can we apply deep learning strategies that self-learn the behavior of a model in order to provide future predictions? See Section \ref{sec:emperical}.
\label{q:1}
\end{question}

\begin{question} \normalfont Can we use real time data as input and apply deep learning strategies instead of models that self learn and provide future predictions? See Section \ref{sec:lstm-covariate}.
\label{q:2}
\end{question}

\begin{question} \normalfont Does the inclusion of covariates improve the prediction quality while including geospatial and socio-economic risk factors? See Section \ref{sec:lstm-covariate}.
\label{q:3}
\end{question}

\begin{question} \normalfont Are there some impact factors that influence the prediction more than others? See Section \ref{sec:lstm-covariate}.
\label{q:4}
\end{question}

\begin{comment}
\begin{question} \normalfont What impact do we see when we introduce many risk factors? See Section \ref{sec:lstm-many}.
\label{q:4}
\end{question}
\end{comment}

\subsection{Deep Learning Predictions of Model-based Empirical Fits} 
\label{sec:emperical}

To establish the feasibility of using LSTM, we have, in our
analysis, verified that through deep learning we can recreate the
predictions identified by a model-based approach using empirical
fits. In this comprehensive analysis, we obtain predictions through
deep learning while using data generated from epidemiological models
as discussed in \citep{marsland20-covid-paper}. The time series we used
was 100 days long, and a multi-layered LSTM
recurrent network was used. Our prediction approach differs by
learning not only from the demographics (fixed data for each city) and
time-dependent data but by integrating the population model for the
underlying prediction, as shown in Figure \ref{fig:magic-1}. Such model
predictive integration capability is important in any application with
multiple time scales. For example, this allows us to integrate
multi-scale time effects into the forecast, which could address the
combination of general forecasts and the next time steps of choice,
such as days, weeks, months, or longer. For this analysis, we used 37
of the 110 cities with reliable empirical fits (that we derived from
an epidemiological model and not deep learning
\citep{marsland20-covid-paper}) for the cases and deaths data up to
April 15, 2020. Our analysis identified a sophisticated single deep
learning time evolution operator that can describe these 37 separate
data sets, and smooth fitted data leads to very accurate deep learning
descriptions via LSTM as depicted in Figure \ref{fig:magic-1}
that are smaller than 1\% from the model predictions.

\begin{figure}[!h]
    \centering

    \begin{subfigure}{.49\textwidth}
      \centering
      \includegraphics[width=1.0\linewidth]{images/predict/Fig4a.pdf}  
      \caption{Prediction acuracy for deaths.}
      \label{fig:magic-a}
    \end{subfigure}
    \begin{subfigure}{0.49\textwidth}
      \centering
      \includegraphics[width=1.0\linewidth]{images/predict/Fig4b.pdf}        
      \caption{Prediction acuracy for cases.}
      \label{fig:magic-b}
    \end{subfigure}

    \caption{Cumulative empirical fits can be very accurately
      predicted with LSTM. The data is taken from 37 U.S. Cities. Our
      comparison showcases that given a model, we can replicate the
      prediction of such models using the model data over time. The
      x-axis represents the days since the first data was fed to the
      model.}
    \label{fig:magic-1}
\end{figure}

\subsection{Realtime-Data Covariate LSTM Prediction}
\label{sec:lstm-covariate}

In this section, we establish that using deep learning can not only be
applied to model predictions but also from real time data. As we want to
evaluate the influence of the risk factors on our prediction framework
using not only models but deep learning algorithms, we have devised a
parameter sweep framework based on LSTM that conducts the prediction
for 110 cities while also considering the risk factors as listed
in Table \ref{tab:risk-factors}. This framework can query any
combination of risk factors as well as not including any, in which
case it defaults to the traditional LSTM algorithm as introduced in
Section \ref{sec:lstm-theory}. Hence, we will answer Questions
\ref{q:2} and \ref{q:3}.

To answer these questions, we chose for our next analysis a date range
between 2020-02-01 and 2020-05-25. We start with a basic LSTM operator
with two layers of LSTM, and initial and final fully connected layers
\citep{Kadupitiya2020-zq}.  For LSTM, we have selected the activation
function based on Rectified Linear Units (RELU). For the last layer
sigmoid is chosen. A drop out rate of 0.2 is used, we use 16 input
nodes, and have a batch size of 110. The  
batch size is the subset size of the training sample (e.g. 110 out of 12540 for a 3 day input length) that is used to train the LSTM during its learning process. The maximum epoch is chosen to be
200. This is the number of steps after we interrupt the learning process. The number of nodes internally is 32. The maximum total number of
samples is 12540 when using 3 days as input length. We use a two-layer
LSTM, as shown in Table \ref{tab:model}. Within our system, we have
the ability to set a number of input days. For this analysis, we have
chosen three different input lengths, namely 3, 4, and 5 days. The
term ``None'' in the output shape means this dimension is
variable.

\begin{table}[!h]
    \caption{The deep learning parameters used for the model sweep
      creation in as exported by Keras using 1 risk factor. The
      activation function is RELU and the recurrent activation
      function is sigmoid.}
    \label{tab:model}
    \bigskip

    \settowidth{\rotheadsize}{Parameters\quad}

    \begin{footnotesize}
    \centering
    \resizebox{1.0\textwidth}{!}{ 
    \begin{tabular}{|ll|lll|ll|ll|}
    \toprule 
     \rothead{Layer} & \rothead{Type}   &
     \rothead{Output Shape} &  \rothead{Parameters for 0 risk factor}  
                  &  \rothead{Parameters for 1 risk factor} &
     \rothead{Output Shape} &  \rothead{Parameters for 33 risk factors} &
     \rothead{Output Shape} &  \rothead{Parameters for 37 risk factors}\\
    \midrule
    dense & Dense    & (None,1,32) & 96   & 128  &  (None,5,32) & 1152 & (None,5,32) & 1280 \\
    LSTM & LSTM      & (None,1,16) & 3136 & 3136 &  (None,5,16) & 3136 & (None,5,16) & 3136 \\
    LSTM 1 & LSTM   & (None,16)  & 2112  & 2112 &  (None,16)   & 2112 & (None,16)   & 2112 \\
    Dense 1 & Dense & (None,32)  &  544  & 544  &  (None,32)   &  544 & (None,32)   &  544 \\
    Dense 2 & Dense & (None,30)  &  990  & 990  &  (None,30)   &  990 & (None,30)   &  990 \\
    \bottomrule
    \end{tabular}
    } ~\\
    The LSTM contains a total of 6,910 trainable parameters in
    case we include one risk factor. In the case, we do not use any
    risk factor it is 6878, and when we include 33 risk factors, it is
    7934.
    \end{footnotesize}
\end{table}

We have run the training on all 110 cities from our data that
represented each FIPS while not including any risk factors (Question
\ref{q:2}) and running them with a single risk factor while iterating
over all risk factors (Question \ref{q:3}). We then summed up all
absolute errors for the predicted data points over the same time
period and repeated each experiment 10 times. We summarized all the
results in the form of a box-whisker diagram each for deaths and for
cases. We sorted the diagram in such a fashion that the risk factor
that has the model with the lowest error appears first. We preceded
the graph with the experiment labeled as ``NONE'' that which does not
use any risk factors to provide a comparison. The graphs are depicted
in Figures \ref{fig:box-cases} for cases, and \ref{fig:box-death} for
deaths.

\begin{figure}[!p]
    \centering
    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_casesNEW2.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the
      prediction of COVID-19 cases. The y-axis represents the
      cumulative error over all input data for the cities.}
    \label{fig:box-cases}
    \bigskip

    \includegraphics[width=1.0\textwidth]{images/boxwhisker/boxplot_deathNEW2.pdf}
    \vspace{-1cm}
    \caption{Influence of risk factors on the accuracy of the
      prediction of COVID-19 death. The y-axis represents the
      cumulative error over all input data for the cities.}
    \label{fig:box-death}
\end{figure}

Further, the obtained sorted order of the best prediction models by
minimum error based on the risk factors for cases and deaths are as
follows:

\begin{description}

\item[Order for cases sorted by minimum model error:] NONE,
  POP\_DENSITY\_2010, PHLTH, INSURANCE, BLACK,
  HISPANICS, NHOSP, CHOLSCREEN, DIABETES, STROKE,
  CHD, CHECKUP, NBEDS, SVI\_OVERALL, KIDNEY, CASTHMA, BPMED,
  LPA, CSMOKING, ARTHRITIS, POVERTY, ESTBEDS, MHLTH, SENIOR, COPD,
  CANCER, NBEDS/1000, BPHIGH, HIGHCHOL, BINGE, OBESITY, SVI\_MINORITY

### 2021-01-27 draft/response-3.tex

\item[Order for deaths sorted by minimum model error:] NONE,
  SVI\_OVERALL, CSMOKING, HISPANICS, POP\_DENSITY\_2010,
  SENIOR, CHOLSCREEN, CASTHMA, INSURANCE, MHLTH, LPA,
  POVERTY, OBESITY, CANCER, BLACK, NBEDS/1000, DIABETES, ARTHRITIS,
  NBEDS, NHOSP, BPHIGH, CHECKUP, PHLTH, CHD, STROKE, HIGHCHOL,
  ESTBEDS, KIDNEY, BINGE, BPMED, COPD, SVI\_MINORITY
\end{description}

We show the errors and the association with a risk factors in Table
 \ref{tab:top-1-cases} and \ref{tab:top-1-deaths}. In the tables we include the following columns (a) the Risk Factor (b) the root mean square error of all models for this risk factor (c) the minimum cumulative error by risk factor and (d) the number of days that wer used as input and identified to relate to the model with the minimum cumulative error. The table has in the first column the rank this particular model comperes to all other models with minimal risk factor. 
The rank was obtained by sorting the best model for each risk factor. 
We display two tables. One for cases (Table \ref{tab:top-1-cases} and one for deaths (Table \ref{tab:top-1-cases}).

We see that in our experiment, almost all risk factors lead to an increased
model prediction accuracy for cases.  We note that by including
factors such as population density, physical health Insurance,
population breakdown by ethnicity, number of hospitals, and diabetes,
lead to better overall predictions for cases. 
We also see that
many of these factors perform better on average.

\begin{table}[!p]
\caption{Best models for cases for each risk factor and their respective
  errors. The model that uses no risk factor is highlighted in grey. We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.}
\label{tab:top-1-cases}
\bigskip
\centering
%\resizebox{1.0\textwidth}{!}{ 
\begin{tabular}{rrrrrlrl}
\toprule
 \makecell{Place/\\Rank} & 
\makecell{Risk Factor} &  
\makecell{RMSE\\Cases} &  
  \makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Cases by\\Risk Factor} & 
 \makecell[r]{Number of \\Days as \\Input} \\
\midrule

1	&	POP\_DENSITY\_2010	&	0.055443	&	7.94	&	5	 \\
2	&	PHLTH	&	0.054795	&	8.01	&	3	 \\
3	&	BLACK	&	0.055028	&	8.12	&	4	 \\
4	&	HISPANIC	&	0.056061	&	8.22	&	5	 \\
5	&	NHOSP	&	0.054621	&	8.26	&	3	 \\
6	&	INSURANCE	&	0.054557	&	8.34	&	3	 \\
7	&	CHOLSCREEN	&	0.054712	&	8.37	&	3	 \\
8	&	DIABETES	&	0.054586	&	8.4	&	5	 \\
9	&	STROKE	&	0.05416	&	8.45	&	4	 \\
10	&	CHD	&	0.054027	&	8.46	&	5	 \\
11	&	CHECKUP	&	0.054927	&	8.51	&	5	 \\
12	&	NBEDS	&	0.054146	&	8.54	&	5	 \\
13	&	SVI\_OVERALL	&	0.054795	&	8.58	&	5	 \\
14	&	KIDNEY	&	0.0546	&	8.61	&	5	 \\
15	&	CASTHMA	&	0.055089	&	8.63	&	5	 \\
16	&	BPMED	&	0.054685	&	8.74	&	3	 \\
17	&	LPA	&	0.054323	&	8.75	&	5	 \\
18	&	CSMOKING	&	0.053716	&	8.75	&	5	 \\
19	&	ARTHRITIS	&	0.054053	&	8.83	&	3	 \\
20	&	POVERTY	&	0.055029	&	8.84	&	3	 \\
21	&	ESTBEDS	&	0.054545	&	8.86	&	4	 \\
22	&	MHLTH	&	0.054864	&	8.9	&	3	 \\
23	&	COPD	&	0.055589	&	8.93	&	4	 \\
24	&	SENIOR	&	0.055079	&	8.93	&	3	 \\
25	&	CANCER	&	0.0545	&	8.99	&	3	 \\
26	&	NBEDS/1000	&	0.054397	&	9.05	&	3	 \\
27	&	BPHIGH	&	0.054609	&	9.13	&	4	 \\
\rowcolor{black!5} 28	&	NONE	&	0.054868	&	9.26	&	3	 \\
29	&	HIGHCHOL	&	0.056429	&	9.28	&	3	 \\
30	&	BINGE	&	0.054317	&	9.32	&	5	 \\
31	&	OBESITY	&	0.054469	&	9.5	&	3	 \\
32	&	SVI\_MINORITY	&	0.055094	&	9.54	&	4	 \\
\bottomrule
\end{tabular}
%}
\end{table}

\begin{table}[!p]
\caption{Best models for deaths for each risk factor and their respective
  errors. The model that uses no risk factor is highlighted in grey.
  We present the Root Mean Square Error (RMSE), the cumulative error and how many days as input were used for this corresponding model.
  }
\label{tab:top-1-deaths}
\bigskip
\centering
%\resizebox{1.0\textwidth}{!}{ 
\begin{tabular}{rrrrrlrl}
\toprule
 \makecell{Place/\\Rank} & 
\makecell{Risk Factor} &  
\makecell{RMSE\\Deaths} &  
\makecell[r]{Minimum\\Cumu-\\lative \\Error for \\Deaths by\\Risk Factor} &  
 \makecell[r]{Number of \\Days as \\Input} \\
\midrule
1	&	SVI\_OVERALL	&	0.067457	&	18.75	&	5	 \\
2	&	CSMOKING	&	0.066898	&	19.01	&	5	 \\
3	&	HISPANIC	&	0.067489	&	19.04	&	5	 \\
\rowcolor{black!5} 4	&	NONE	&	0.06696	&	19.09	&	4	 \\
5	&	POP\_DENSITY\_2010	&	0.066968	&	19.13	&	4	 \\
6	&	SENIOR	&	0.068498	&	19.27	&	3	 \\
7	&	CHOLSCREEN	&	0.067859	&	19.43	&	5	 \\
8	&	CASTHMA	&	0.066952	&	19.46	&	5	 \\
9	&	INSURANCE	&	0.067127	&	19.48	&	5	 \\
10	&	MHLTH	&	0.068742	&	19.53	&	3	 \\
11	&	LPA	&	0.067149	&	19.54	&	5	 \\
12	&	POVERTY	&	0.067267	&	19.57	&	5	 \\
13	&	OBESITY	&	0.067123	&	19.68	&	5	 \\
14	&	CANCER	&	0.068363	&	19.70	&	3	 \\
15	&	NBEDS/1000	&	0.06752	&	19.73	&	5	 \\
16	&	BLACK	&	0.068352	&	19.83	&	3	 \\
17	&	DIABETES	&	0.067636	&	19.84	&	5	 \\
18	&	ARTHRITIS	&	0.066927	&	19.94	&	4	 \\
19	&	NBEDS	&	0.066879	&	19.95	&	4	 \\
20	&	NHOSP	&	0.067391	&	19.95	&	5	 \\
21	&	BPHIGH	&	0.067799	&	19.99	&	5	 \\
22	&	PHLTH	&	0.068737	&	20.02	&	3	 \\
23	&	CHECKUP	&	0.067543	&	20.02	&	5	 \\
24	&	CHD	&	0.06874	&	20.15	&	3	 \\
25	&	STROKE	&	0.066906	&	20.16	&	4	 \\
26	&	HIGHCHOL	&	0.068557	&	20.35	&	3	 \\
27	&	ESTBEDS	&	0.067052	&	20.42	&	4	 \\
28	&	KIDNEY	&	0.068948	&	20.49	&	3	 \\
29	&	BINGE	&	0.067447	&	20.56	&	5	 \\
30	&	BPMED	&	0.067399	&	20.69	&	5	 \\
31	&	COPD	&	0.06745	&	20.98	&	5	 \\
32	&	SVI\_MINORITY	&	0.067443	&	21.53	&	4	 \\
\bottomrule
\end{tabular}
%}
\end{table}

We observed a different scenario for deaths, in which we only
identified two risk factors, namely the Social Vulnerability Index
(SVI) and chain-smoking, that, when integrated into our deep learning
model, leads to better prediction results. However, the accuracy of
the overall prediction of deaths is far less precise than that of the
cases. To showcase this, we have included Figures
\ref{fig:error-case-popdensity} and \ref{fig:error-death-popdensity}
that show the respective errors.

\begin{figure}[!h]
  \begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Cases.pdf}
    %\vspace{-1cm}
    \caption{Model prediction error cases.}
    \label{fig:error-case-popdensity}
  \end{subfigure}
  \begin{subfigure}{.49\textwidth}
    \centering
    \includegraphics[width=1.0\textwidth]{images/predict/Length_5-fields_pop_density_2010-Cumulative-Death.pdf}
    %\vspace{-1cm}
    \caption{Model prediction error deaths.}
    \label{fig:error-death-popdensity}
    \end{subfigure}

    \caption{Model prediction error for cumulative deaths when
      including one risk factor POP\_DENSITY\_2010
      in addition to the past number of cases and deaths.}
    \label{fig:error-death-popdensity-both}

\end{figure}

This can be explained by our input data distribution between cases and deaths in which we noted higher
fluctuations relative to the overall value in cases of deaths. 
A mitigation to this issue
would be using a seven day average over the analyzed period. However,
we have not done this on purpose for this analysis as we wanted to
identify how a covariate enhanced LSTM behaves that includes risk
factors based on the daily fluctuations. This was done to avoid and
showcase any needed preproccessing on the data and to identify the
capability of the deep learning framework while adapting to the
fluctuations without any special activities. Previously, we have
already shown in Section \ref{sec:emperical} that for smooth data
inputs the prediction is very accurate.  Using fluctuation data allows
our framework for an automated ingest of data on a daily basis to
enable data fusion of new cases and death information. Hence, we can
re-train the model with new incoming data every day. Through repeated
training experiments, we found the LSTM hyperparameters such as epochs
and dropout values that work well for our experiments.

We conducted an additional analysis and plotted the top 10 best
cumulative prediction model errors for each risk factor in
Figure \label{fig:place-top10} for cases and deaths.  The horizontal
black line shows the first occurrence of no risk factor being used.

Notably, we identified that a jump occurs when we sort the top 10
model predictions at around 330 models from a total of 3400
automatically generated models. This also underscores the need to run
our deep learning model multiple times to obtain best parameter
candidates.  Furthermore, we see that the cumulative error for cases
is about half the size as for death. This is caused by cases having a
larger number of samples than those from deaths.

% In case of cases we also see a drop in the better
% model predictions.

\begin{figure}[!htbp]
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for cases over all risk factors.}
%        \label{fig:place-top1-cases}
%
%    \end{minipage}
%    \ \
%    \begin{minipage}{0.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop1_Death.pdf}
%        \vspace{-1cm}
%        \caption{Best predictions for deaths over all risk factors.}
%        \label{fig:place-top1-death}
%    \end{minipage}

%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Cases.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all cases over all risk factors.}
%        \label{fig:place-top5-cases}
%    \end{minipage}
%    \ \
%    \begin{minipage}{.45\textwidth}
%        \centering
%        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop5_Death.pdf}
%        \vspace{-1cm}
%        \caption{Top 5 predictions from all deaths  over all risk factors.}
%        \label{fig:place-top5-death}
%    \end{minipage}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Death.pdf}
        \caption{Cumulative error for deaths.}
        \label{fig:place-top10-death}
    \end{subfigure}
    \begin{subfigure}{.95\textwidth}
        \centering
        \includegraphics[width=1.0\textwidth]{images/predict/PlaceTop10_Cases.pdf}
        \caption{Cumulative error for cases}
        \label{fig:place-top10-cases}
    \end{subfigure}

    \caption{Top 10 predictions from all different risk factors sorted by cumulative error for deaths (\ref{fig:place-top10-death}) and cases (\ref{fig:place-top10-cases}). The horizontal line in each figure represents the model cumulative error with no risk factor as input for deaths and cases respectively.}
    \label{fig:place-top10-both}

\end{figure}

\begin{comment}

\subsection{Impact and Design of a Multi-Covariate-LSTM}
\label{sec:lstm-many}

In our next comprehensive analysis, we try to answer Question
\ref{q:4} if all risk factors impact the analysis. Thus we feed the
covariates into the recurrent network as additional features. In
addition to cases and deaths, this leads to 35 variables, including 33
unique covariates. We looked at two related ways of doing this: (a)
have covariates fixed in time (static), and (b) multiply the
covariates by the times series (we used the average of cases and
deaths)

We expected the second choice to be the most natural as it gives all
35 input streams similar time dependence. However, in our initial
tests, we found these choices gave very similar results, and so we
focused on the simpler choice (a) in this paper. In the experiments
discussed here, we either used 0, 1, or 33 covariates. Choosing 33
gives the best description (reducing loss by about 1\%) while using
each covariate as a solitary extra feature allows us a rather
sensitive study of the importance of each individual covariate as we
noted in Section \ref{sec:lstm-covariate}.

The final choice was the quantities to predict and train on. We choose
this to be both the cases and deaths values at the time value
following the sequence combined with the same quantities for the
following 14 days. This gives us 30 different output features for each
sequence. We renormalized, so the two basic features (following day)
contributed 50\% of the loss function. We supported this in TensorFlow
with a custom mean square error (MSE) loss function that dropped
training values that were not available for time sequences near the
end of the examined period. While one can not train on future values,
we can predict them, as seen in our sample outputs. We extensively
studied both 2 and 30 output features, and they give similar
descriptions of the observed data with the 30 feature choice enabling
14-day predictions.

We performed an extensive hyper-parameter search and we show the
values in Table \ref{tab:model} that give robust results without
overfitting. Furthermore, we looked at four different activation
functions for this analysis in the LSTM and fully connected
layers. Scaled Exponential Linear Unit (SELU), REctified Linear Unit
(RELU) and TANH activations gave very similar results while SIGMOID
activation was much worse. However we still use a SIGMOID function in
the last stage just as we did in Section \ref{sec:lstm-covariate}. We
also looked at the number of hidden units (separately for LSTM and
fully connected layers), the number of LSTM layers and the dropout
values.  Only the latter produced a significant change and, for
example, removing dropouts decreased the loss by about 16\% but we
retained them to provide a robust environment. We chose randomly 20\%
of the data for validation and testing and this had similar loss
values to the training data (the other 80\%) with our chosen dropout.

The final technical issue concerns the data normalization where each
feature is transformed by a feature dependent -- but city independent
-- way to lie between 0 and 1. Further all extensive (proportional to
size) covariates were first divided by the city population except for
the population itself where we took the logarithm; intensive
covariates were just rescaled to lie between 0 and 1. The basic
COVID-19 daily data is extensive but compared two approaches with and
without renormalization, but in this case, we took the square root of
the data value.  This is natural if there are counting (square root(N)
for N observations) errors as the deep learning loss function is
$(predicted-observed)^{2}$ without the error traditional in
$\chi^{2}$, namely $((predicted-observed)/error)^{2}$. This approach
with standard deep learning MSE appears sound instead of dividing by
the population. That will emphasize the error contribution of smaller
cities but is still interesting to look at due to the presence of many
systematic errors.

\section{Direct Deductions from Deep Learning Data Representation}

The deep learning model contains a lot of information much of which we
have already analyzed and discussed in Section
\ref{sec:lstm-covariate} and Figures \ref{fig:magic-1} and
\ref{fig:error-case-popdensity}. In addition, we have looked at the
sensitivity of the results as measured by the predicted daily rates
and the value of the loss function. In Table \ref{tab:corr-matrix}, we
present one way of looking at it. We calculate the correlation
coefficients of many of the covariates from an accurate LSTM model fit
to the data. We divide the time series into three regions

\begin{itemize}
\item \textbf{Past} or all days up to 14 days before the final
  observed date (May 25, 2020)
\item \textbf{Now} or the 14 days from May 12 to May 25, 2020
\item \textbf{Future} or the 14 days after May 25, 2020 predicted by
  the model
\end{itemize}

These results are different between population normalized or direct
non-normalized data. We have kept the square-root used in the model
but divided the model values by the square root of the population when
computing the correlation. Significant positive correlations are seen
for CASTHMA, HIGHCHOL, DIABETES, OBESITY, CSMOKING, CHD, and CHECKUP ,
while negative correlations for PVI and NORM\_POP. We also looked at
the impact of these covariates on the evolution function. As this
operates on data at previous time values, this is sensitive to
variations across cities and times. The evolution operator is
intuitively the time derivative of the absolute values of daily rates
seen in the correlation in Table 3. Nhosp and Estbeds have the most
significance in impacting (as a reduction) the values of both the
infection and fatality rates from the evolution operator. The fatality
rates are also significantly impacted by INSURANCE and NORM\_POP
(reduction) and SENIOR and OBESITY (increase) from the evolution
operator. NORM\_POP impacts infection values as an increase.  We can
also identify which covariates have the most significant impact on the
loss function. Here NHOSP, CHD, KIDNEY, BINGE, and PHLTH give 5\% or
more reduction in loss function while NORM\_POP, SVI\_OVERALL,
CASTHMA, DIABETES, and BPHIGH have small (1\%) effects.

\begin{table}[!h]
\caption{Correlation Matrix}
\bigskip
\label{tab:corr-matrix}
\begin{center}
\begin{tabular}{lrrrrrr}
\toprule
Risk Factor		&
\makecell{Past\\Case} &
\makecell{Now\\Case} &
